---
title: "The bnbc User's Guide"
shorttitle: "bnbc guide"
author: 
  - Kipper Fletez-Brant
  - Kasper Daniel Hansen
package: bnbc
bibliography: bnbc.bib
abstract: >
  A comprehensive guide to using the bnbc package for normalizing Hi-C replicates.
vignette: >
  %\VignetteIndexEntry{bnbc User's Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output: 
  BiocStyle::html_document:
    toc_float: true
---

# Introduction

The `r Biocpkg("bnbc")` package provides functionality to perform normalization and batch correction **across samples** on data obtained from Hi-C [@HiCassay] experiments.

In this package we implement tools for general subsetting of and data extraction from sets of Hi-C contact matrices, as well as smoothing of contact matrices, cross-sample normalization and cross-sample batch effect correction methods.

`r Biocpkg("bnbc")` expects as input 

1. a `GRanges` object representing the genome assayed, with individual ranges having the width that is equal to the bin size used to partition the genome. 
2. a `list` of (square) contact matrices.
3. a `DataFrame` object containing sample-level covariates (i.e. gender, date of processing, etc).

## Citing bnbc

Please cite our bioRxiv paper.

## Terminology

It is well appreciated that Hi-C contact matrices exhibit an exponential decay in observed number of contacts as a function of the distance between the pair of interacting loci. In this work we operate, as has recently been done (i.e. [@HiCRep]), on the set of all loci interacting at a specific distance, one chromosome at a time. For a given distance $k$, the relevant set of loci are listed in each contact matrix as the entries comprising the $k$-th matrix diagonal (with the main diagonal being referred to as the first diagonal). We refer to these diagonals as matrix "bands".

# Dependencies

This document has the following dependencies

```{r dependencies, warning=FALSE, message=FALSE}
library(bnbc)
```

# bnbc class

`r Biocpkg("bnbc")` uses the `ContactGroup` class to represent the set of contact matrices for a given set of genomic locis interactions. The class has 3 slots:

- `rowData`: a `GRanges` object that has 1 element for each bin of the partitioned genome.
- `colData`: a `DataFrame` object that contains information on each sample (i.e. gender).
- `contacts`: a `list` of contact matrices.

We expect each `ContactGroup` to represent data from 1 chromosome. We are thinking about a whole-genome container.  An example dataset for chromosome 22 is supplied with the package.

```{r dataLoad}
data(cgEx)
cgEx
```

Creating a `ContactGroup` object requires specifying the 3 slots above:

```{r create}
cgExCreate <- ContactGroup(rowData=rowData(cgEx),
                           contacts=contacts(cgEx),
                           colData=colData(cgEx))
```

Note that in this example, we used the accessor methods for each of these slots; there are also corresponding 'setter' methods, such as `rowData(cgEx)<-`. 

Printing a `ContactGroup` object gives the number of bins represented by the `rowData` slot, the width of the bin in terms of genomic distances (i.e. 100kb) and the number of samples:

```{r print}
cgEx
```

## Getting your data into bnbc

What is needed is essentially a list of contact matrices. We assume that the contact matrices are square, with no missing values. We do not require that data have been transformed or pre-processed by various bias correction software and provide methods for some simple pre-processing techniques.

Below, we demonstrate the steps needed to convert a set of hypothetical contact matrices into a `ContactGroup` object. The object `upper.mats.list` is supposed to be a list of contact matrices, each represented as an upper triangular matrix. We also suppose `LociData` to be a `GenomicRanges` object containing the loci of the contact matrices, and `SampleData` to be a `DataFrame` of per-sample covariates (i.e. cell type, sample name etc). We first convert all contact matrices to be symmetric matrices, then use the constructor method `ContactGroup()` to create the object.

```{r data_to_bnbc, eval=FALSE, echo=TRUE}
## Example not run
## Convert upper triangles to symmetry matrix
MatsList <- lapply(upper.mats.list, function(M) M[lower.tri(M)] = M[upper.tri(M)])
## Use ContactGroup constructor method
cg <- ContactGroup(LociData, MatsList, SampleData)
```

# Working with bnbc contact matrices

We provide setter and getter methods for manipulating individual matrix bands for contact matrices as well.  First, we have functions for working with bands of individual matrices (not bnbc related):

```{r band_example}
mat.1 <- contacts(cgEx)[[1]]
mat.1[1000:1005, 1000:1005]
b1 <- band(mat=mat.1, band.no=2)
band(mat=mat.1, band.no=2) <- b1 + 1
mat.1[1000:1005, 1000:1005]
```

In this example, the main diagonal of the contact matrix is also the main diagonal of the printed example above.  Similarly, band number two, which is also the first off-diagonal, is also the first off-diagonal of the printed example.  As can be seen from the printed example, updating a matrix band is a symmetric operation, and updated the first off-diagonal in both the upper and lower triangles of the matrix.

To utilize this across a list of contact matrices, we have the `cgApply()` function which applies the same function to each of the matrices. It supports parallelization using `r CRANpkg("parallel")`.

# Per-Sample Adjustments

To adjust for differences in depth sequencing, we first apply the `logCPM` transform [@voom] to each contact matrix.  This transformation divides each contact matrix by the sum of the upper triangle of that matrix (adding 0.5 to each matrix cell and 1 to sum of the upper triangle), scales the resulting matrix by $10^6$ and finally takes the log of the scaled matrix (a fudge factor is added to both numerator and denominator prior to taking the logarithm)..

```{r logcpm}
cgEx.cpm <- logCPM(cgEx)
```

Additionally, we smooth each contact matrix with a square smoothing kernel to reduce artifacts of the choice of bin width.  We support both box and Gaussian smoothers.

```{r smoothing}
cgEx.smooth <- boxSmoother(cgEx.cpm, h=5)
## or
## cgEx.smooth <- gaussSmoother(cgEx.cpm, radius=3, sigma=4)
```

# Cross Sample Normalization

BNBC operates on each matrix band separately.  For each matrix band $k$, we extract each sample's observation on that band and form a matrix $M$ from those bands;  if band $k$ has $d$ entries, then after `logCPM` transformation,  $M \in \mathbb{R}^{n \times d}$.  For each such matrix, we first apply quantile normalization [@QN] to correct for distributional differences, and then ComBat [@ComBat] to correct for batch effects.

Here we will use `r Biocpkg("bnbc")` to do batch correction on the first 10 matrix bands, beginning with the second matrix band and ending on the eleventh.

```{r bnbc}
cgEx.bnbc <- bnbc(cgEx.smooth, batch=colData(cgEx.smooth)$Batch,
                  threshold=1e7, step=4e4, nbands=11)
```

# sessionInfo()

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References


